### 2014-10-30 submission_v1-10-29-14/Paper_v2.tex
Submission on 10-29-14

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Reduced order modeling of blood coagulation}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Processes}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
In this study, we present a novel modeling approach which combines ordinary differential equation (ODE) 
modeling with logical rules to simulate an archetype biochemical network, the human coagulation cascade.  
The model consisted of five differential equations augmented with several logical rules describing 
regulatory connections between model components, and unmodeled interactions in the network. 
This formulation was more than an order of magnitude smaller than current coagulation models, 
because many of the mechanistic details of coagulation were encoded as logical rules.
We estimated an ensemble of likely model parameters (N = 20) from \textit{in~vitro} extrinsic coagulation data sets, with and without inhibitors, 
by minimizing the residual between model simulations and experimental measurements using particle swarm optimization (PSO). 
We then validated the model ensemble using thrombin data sets that were not used during training. 
The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
in the presence of platelets (a significant unmodeled component).
We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
performance of the network. Taken together, the hybrid approach produced a surprisingly predictive model given its small size, 
suggesting the proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
or potentially even cell free metabolic systems.

@@

Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
using mass action or more complex kinetics, to describe the rates of biochemical conversions \citep{Khanin:1989aa,Willems:1991aa,Baldwin:1994aa,Leipold:1995aa,Kuharsky:2001aa}.
Mechanistic ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys} 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, Hockin \emph{et al.} \citep{Hockin:2002aa}, and later Butenas et al., \citep{Butenas:2004aa}
who developed and then subsequently refined highly mechanistic models of coagulation. 
Other laboratories have also expanded upon Hockin \emph{et al.}, for example by exploring the intrinsic pathway, the role of stochastic fluctuations in coagulation \citep{Lo:2005aa},
and the dynamics of thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside clots \citep{Voronov:2013aa}.
However, these previous studies were largely based upon extensive mechanistic knowledge. 
This is possible because blood, while enormously complex, can be systematically interrogated.
Other systems, such as intracellular signaling networks, are much more difficult to experimentally interrogate.  
Towards this unmet need, we formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which a complete mechanistic understanding is missing. 
We tested this approach by modeling the human coagulation cascade. 
The hybrid model consisted of only five differential equations augmented with several logical rules.
Thus, the model was more than an order of magnitude smaller than comparable purely ODE models in the literature.
We estimated the model parameters from \textit{in~vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then compared the model predictions with thrombin data sets, for both normal and hemophilic coagulation, that were not used for model training.
Once validated, we performed flux and sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, 
suggesting this framework could potentially be used to model other biochemical networks important to human health.

@@

The model ensemble predicted a direct correlation between thrombin generation and rFVIIa addition in hemophilia (Fig. \ref{fig-rFVII}).
In the current model, we cannot distinguish between different initiation sources, e.g., TF/FVIIa versus rFVIIa, as we have only a single lumped initiation source (trigger).
Thus, we simulated the addition of rFVIIa in hemophilia by removing fVIII and fIX from the model, and modulating the initial level of trigger.
Simulations with a baseline level of trigger were consistent with the previous hemophilia simulations, where the only thrombin produced was from initiation (Fig. \ref{fig-rFVII}A, 1$\times$~trigger).
However, as we increased the trigger strength, the thrombin profile began to approximate normal coagulation, showing a pronounced peak albeit with a slower peak time (Fig. \ref{fig-rFVII}B, $t^{**}>t^{*}$).
Further increases in trigger strength resulted in decreased thrombin peak time and increased maximum thrombin values (Fig. \ref{fig-rFVII}A, 50$\times$~trigger). 
Thus, for large trigger values (200$\times$trigger), the hemophilic thrombin profile approximated normal coagulation, where peak thrombin was achieved shortly 
after administration and 95\% of the thrombin was gone by 20 min after initiation. 
We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and inhibition of thrombin formation for normal coagulation and hemophilia.
Analysis of the reaction flux through the reduced order network for thrombin generation in normal, hemophilia and rFVIIa-treated hemophilia identified three distinct operational modes (Fig. \ref{fig-flux-analysis}).
We calculated the flux through four lumped reactions, initiation, amplification, thrombin-induced APC generation and total thrombin inhibition (including both APC and ATIII action).
Directly after the addition of a trigger (e.g., TF/FVIIa or rFVIIa), the lumped initiation flux was the largest for all three cases.
However, within a few minutes enough thrombin was generated by the initiation mechanism to induce the amplification stage. 
During amplification, thrombin catalyzes its own formation and inhibition by generating activated protein C (APC), a potent inhibitor of the coagulation cascade.
For normal coagulation, amplification and thrombin inhibition were the dominate reactions by 6 min after initiation (Fig. \ref{fig-flux-analysis}, left). 
After 10 min, the dominate reaction had shifted to thrombin inhibition (both ATIII and APC action). 
In hemophilia (missing both fVIII and fIX), the amplification reaction did not occur, and thrombin was produced only by initiation (Fig. \ref{fig-flux-analysis}, center).
Initiation was quickly inhibited by APC, and the thrombin level stabilized (eventually decaying at longer times because of ATIII activity). 
Lastly, when 50$\times$trigger was used to induce thrombin formation in hemophilia (absence of fVIII/fIX), 
initiation mechanisms dominated for up to 6 min following initiation (Fig. \ref{fig-flux-analysis}, right). 
Similar to hemophilia alone, no amplification occurred in the 50$\times$trigger+hemophilia case, 
and the rate of thrombin generation was extinguished by the combined action of ATIII and APC. 
Taken together, the hybrid modeling approach captured the transition between the modes of thrombin generation, as well as the role that inhibitors play 
in attenuating the thrombin generation rate. Thus, the transfer function approach encoded the inhibitory logic of this cascade in the absence of specific mechanism.

@@

\section*{Discussion}
In this study, we developed a reduced order model of the human coagulation cascade.
We modeled coagulation because it is well studied, has a complex architecture, 
and has an abundance of experimental data available for model identification and validation. 
However, coagulation was just a proof of concept test of our approach. 
The proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
or potentially even cell free metabolic systems. 
The model consisted of five differential equations augmented with several logical rules describing 
regulatory connections between model components and unmodeled interactions in the network. 
We estimated model parameters from \textit{in vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
To estimate parameters, the residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
However, not all of the model parameters were uniquely identifiable, given the training data.
Instead, we estimated an ensemble of likely parameter sets (N = 20)
from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway.
Ensemble approaches have been used previously for other signal transduction models \citep{Kuepfer:2007aa,Song:2009aa,2010_song_varner_BiotechJ,Tasseff:2010aa,Lequieu:2011aa}, 
and for metabolic models \citep{Tran:2008aa} to 
estimate the impact of poorly constrained parameter values or poorly understood network structure on simulation performance. 
Thus, ensemble approaches are common in the dynamic modeling community. 
However, perhaps a unique feature of the current study is the direct connection between our particle swarm approach, 
and the parameter ensemble; each particle in our swarm uniquely corresponded to a parameter set in our ensemble. 
Thus, by constraining particles to operate in different parameter regions, giving each particle a different parameter combination to explore, 
or perhaps even suppling a different model formulation to each particle we can effectively 
traverse through complex parameter and model spaces. 
We validated the ensemble using thrombin data sets taken from multiple laboratories for a variety of experimental conditions not used during training. 
The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
in the presence of platelets (a significant unmodeled component).
We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
performance of the network. Taken together, the proposed hybrid framework produced a surprisingly predictive model, suggesting this approach could 
be used to effectively model other biochemical networks important to human health.

@@

Although the performance of the proof of principle coagulation model was impressive given its limited size, there were several issues that could be further explored.
First, the prediction of initiation time should be investigated. 
We were able to estimate initiation time within a data set, but unable to predict initiation time \textit{across} independent data sets. 
This suggested that we should update the initiation module to distinguish between different triggers, e.g., TF/FVIIa versus rFVIIa alone, 
and to include key biological milestones such as FXa activation (a prerequisite to thrombin formation).
Next, there are several additional biological modules that could be added to the core model presented here. 
First, we could include thrombin-induced platelet activation and the role of activated platelets in amplification.
We captured thrombin generation data in the presence of platelets, however,
the initial shape of the activation curve and the time-scale of activation was not always consistent with the data.
Platelets are activated by thrombin through the cleavage of the extracellular domain of protease-activated receptors (PARs) on the platelet surface.
Once activated, platelets play an important role in amplification, and are key mediators of the positive feedback driving amplification.
Thus, this biology is a potentially important component of an expanded model.
We should also add the intrinsic pathway to the model. 
The intrinsic pathway is triggered by contact activation of the plasma protease factor XI (fXI) by negatively charged surfaces and by thrombin and upstream factors such as activated
plasma protease factor XII (FXIIa) \citep{NAITO1991,Gailani:1991aa}.
Activated platelets may also release polyphosphate which directly activates fXII \cite{Smith:2006zl}. 
Arguably a minor player in acute bleeding, contact activation could also be important in other wound healing contexts. 
Finally, to make the model more clinically relevant, we should include the biochemical processes responsible for clot formation and clot dissolution (fibrinolysis).
Clot formation is driven by thrombin activity, while fibrinolysis is driven by plasmin activity, a key enzyme that cleaves fibrin (one of the main materials in a clot).
Similar to coagulation, fibrinolysis is managed by several activating and inhibitory factors which control the balance between clot formation and dissolution.
Tissue plasminogen activator (t-PA) and urokinase activate plasmin, along with contact pathway factors such as fXIa. On the other hand, thrombin activatable fibrinolysis inhibitor (TAFI)
inhibits the degradation of fibrin by plasmin. Also, similar to coagulation, there is considerable fibrinolysis and contact pathway data sets that can be used to train the model.
Lastly, the choice of max/min integration rules or the particular form of the transfer functions could be generalized to include other rule types and functions. 
Theoretically, an integration rule is a function whose domain is a set of transfer function inputs, and whose range is $v\in[0,1]$.
Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
Alternative integration rules such as the mean might have different properties which could influence model identification or performance. 
For example, a mean integration rule would be differentiable, which allows derivative-based optimization approaches to be used. 
The particular form of the transfer function could also be explored. We choose a Hill-like function because of its 
prominence in the systems and synthetic biology community. 
However, the only mathematical requirement for a transfer function is that it map a non-negative continuous or categorical variable into the 
range $f\in[0,1]$. Thus, many transfer function types are possible.

>>>>

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Reduced order modeling of blood coagulation}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Processes}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
In this study, we present a novel modeling approach which combines ordinary differential equation (ODE) 
modeling with logical rules to simulate an archetype biochemical network, the human coagulation cascade.  
The model consisted of five differential equations augmented with several logical rules describing 
regulatory connections between model components, and unmodeled interactions in the network. 
This formulation was more than an order of magnitude smaller than current coagulation models, 
because many of the mechanistic details of coagulation were encoded as logical rules.
We estimated an ensemble of likely model parameters (N = 20) from \textit{in~vitro} extrinsic coagulation data sets, with and without inhibitors, 
by minimizing the residual between model simulations and experimental measurements using particle swarm optimization (PSO). 
Each parameter set in our ensemble corresponded to a unique particle in the PSO.
We then validated the model ensemble using thrombin data sets that were not used during training. 
The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
in the presence of platelets (a significant unmodeled component).
We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
performance of the network. Taken together, the hybrid approach produced a surprisingly predictive model given its small size, 
suggesting the proposed framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
or potentially even cell free metabolic systems.

@@

Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
using mass action or more complex kinetics, to describe the rates of biochemical conversions \citep{Khanin:1989aa,Willems:1991aa,Baldwin:1994aa,Leipold:1995aa,Kuharsky:2001aa}.
Mechanistic ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys} 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, Hockin \emph{et al.} \citep{Hockin:2002aa}, and later Butenas et al., \citep{Butenas:2004aa}
who developed and then subsequently refined highly mechanistic coagulation models. 
Other laboratories have also expanded upon Hockin \emph{et al.}, for example by exploring the intrinsic pathway, the role of stochastic fluctuations in coagulation \citep{Lo:2005aa},
and the dynamics of thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside clots \citep{Voronov:2013aa}.
However, these previous studies were largely based upon extensive mechanistic knowledge. 
This is possible because blood, while enormously complex, can be systematically interrogated.
Other systems, such as intracellular signaling networks, are much more difficult to experimentally interrogate.  
Towards this unmet need, we formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which a complete mechanistic understanding is missing. 
We tested this approach by modeling the human coagulation cascade. 
The hybrid model consisted of only five differential equations augmented with several logical rules.
Thus, the model was more than an order of magnitude smaller than comparable purely ODE models in the literature.
We estimated the model parameters from \textit{in~vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then compared the model predictions with thrombin data sets, for both normal and hemophilic coagulation, that were not used for model training.
Once validated, we performed flux and sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, 
suggesting this framework could potentially be used to model other biochemical networks important to human health.

@@

The model ensemble predicted a direct correlation between thrombin generation and rFVIIa addition in hemophilia (Fig. \ref{fig-rFVII}).
In the current model, we cannot distinguish between different initiation sources, e.g., TF/FVIIa versus rFVIIa, as we have only a single lumped initiation source (trigger).
Thus, we simulated the addition of rFVIIa in hemophilia by removing fVIII and fIX from the model, and modulating the initial level of trigger.
Simulations with a baseline level of trigger were consistent with the previous hemophilia simulations, where the only thrombin produced was from initiation (Fig. \ref{fig-rFVII}A, 1$\times$~trigger).
However, as we increased the trigger strength, the thrombin profile began to approximate normal coagulation, showing a pronounced peak albeit with a slower peak time (Fig. \ref{fig-rFVII}B, $t^{**}>t^{*}$).
Further increases in trigger strength resulted in decreased thrombin peak time and increased maximum thrombin values (Fig. \ref{fig-rFVII}A, 50$\times$~trigger). 
Thus, for large trigger values (200$\times$trigger), the hemophilic thrombin profile approximated normal coagulation, where peak thrombin was achieved shortly 
after administration and 95\% of the thrombin was gone by 20 min after initiation. 
We performed flux analysis to understand how the reduced order coagulation model balanced initiation, amplification and termination of thrombin generation for normal and hemophilic coagulation.
Analysis of the reaction flux through the reduced order network for thrombin generation in normal, hemophilia and rFVIIa-treated hemophilia identified three distinct operational modes (Fig. \ref{fig-flux-analysis}).
We calculated the flux through four lumped reactions, initiation, amplification, thrombin-induced APC generation and total thrombin inhibition (including both APC and ATIII action).
Directly after the addition of a trigger (e.g., TF/FVIIa or rFVIIa), the lumped initiation flux was the largest for all three cases.
However, within a few minutes enough thrombin was generated by the initiation mechanism to induce the amplification stage. 
During amplification, thrombin catalyzes its own formation and inhibition by generating activated protein C (APC), a potent inhibitor of the coagulation cascade.
For normal coagulation, amplification and thrombin inhibition were the dominate reactions by 6 min after initiation (Fig. \ref{fig-flux-analysis}, left). 
After 10 min, the dominate reaction had shifted to thrombin inhibition (both ATIII and APC action). 
In hemophilia (missing both fVIII and fIX), the amplification reaction did not occur, and thrombin was produced only by initiation (Fig. \ref{fig-flux-analysis}, center).
Initiation was quickly inhibited by APC, and the thrombin level stabilized (eventually decaying at longer times because of ATIII activity). 
Lastly, when 50$\times$trigger was used to induce thrombin formation in hemophilia (absence of fVIII/fIX), 
initiation mechanisms dominated for up to 6 min following initiation (Fig. \ref{fig-flux-analysis}, right). 
Similar to hemophilia alone, no amplification occurred in the 50$\times$trigger+hemophilia case, 
and the rate of thrombin generation was extinguished by the combined action of ATIII and APC. 
Taken together, the hybrid modeling approach captured the transition between the modes of thrombin generation, as well as the role that inhibitors play 
in attenuating the thrombin generation rate. Thus, the transfer function approach encoded the inhibitory logic of this cascade in the absence of specific mechanism.

@@

\section*{Discussion}
In this study, we developed a reduced order model of the human coagulation cascade.
We modeled coagulation because it is well studied, has a complex architecture, 
and has an abundance of experimental data available for model identification and validation. 
However, coagulation was just a proof of concept test of our approach. 
The proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
or potentially even cell free metabolic systems. 
The model consisted of five differential equations augmented with several logical rules describing 
regulatory connections between model components and unmodeled interactions in the network. 
We estimated model parameters from \textit{in vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
To estimate parameters, the residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
However, not all of the model parameters were uniquely identifiable, given the training data.
Instead, we estimated an ensemble of likely parameter sets (N = 20)
from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway.
Ensemble approaches have been used previously for other signal transduction models \citep{Kuepfer:2007aa,Song:2009aa,2010_song_varner_BiotechJ,Tasseff:2010aa,Lequieu:2011aa}, 
and for metabolic models \citep{Tran:2008aa} to 
estimate the impact of poorly constrained parameter values or poorly understood network structure on simulation performance. 
Thus, ensemble approaches are common in the dynamic modeling community. 
However, a unique feature of the current study is the direct connection between our particle swarm approach, 
and the parameter ensemble; each particle in our swarm uniquely corresponded to a parameter set in our ensemble. 
Thus, by constraining particles to operate in different parameter regions, giving each particle a different parameter combination to explore, 
or perhaps even suppling a different model formulation to each particle we can effectively 
traverse through complex parameter and model spaces. 
We validated the ensemble using thrombin data sets taken from multiple laboratories for a variety of experimental conditions not used during training. 
The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
in the presence of platelets (a significant unmodeled component).
We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
performance of the network. Flux analysis showed the logical rules formulation encoded the transitions between initiation, amplification and termination of thrombin generation.
Sensitivity analysis suggested that the amplification rate constant was more important to the time to peak thrombin, while the ATIII inhibition constant controlled thrombin exposure.
Taken together, the proposed hybrid framework produced a surprisingly predictive model, suggesting this approach could 
be used to effectively model other biochemical networks important to human health.

@@

The performance of the proof of principle coagulation model was impressive given its limited size, however, there are several issues that could be further explored.
First, the prediction of initiation time should be investigated. 
We were able to estimate initiation time within a data set, but unable to predict initiation time \textit{across} independent data sets. 
This suggested that we should update the initiation module to distinguish between different triggers, e.g., TF/FVIIa versus rFVIIa alone, 
and to include key biological milestones such as FXa activation (a prerequisite to thrombin formation).
Next, there are several additional biological modules that could be added to the core model presented here. 
First, we could include thrombin-induced platelet activation and the role of activated platelets in amplification.
We captured thrombin generation data in the presence of platelets, however,
the initial shape of the activation curve and the time-scale of activation was not always consistent with the data.
Platelets are activated by thrombin through the cleavage of the extracellular domain of protease-activated receptors (PARs) on the platelet surface.
Once activated, platelets play an important role in amplification, and are key mediators of the positive feedback driving amplification.
Thus, this biology is a potentially important component of an expanded model.
We should also add the intrinsic pathway to the model. 
The intrinsic pathway is triggered by contact activation of the plasma protease factor XI (fXI) by negatively charged surfaces and by thrombin and upstream factors such as activated
plasma protease factor XII (FXIIa) \citep{NAITO1991,Gailani:1991aa}.
Activated platelets may also release polyphosphate which directly activates fXII \cite{Smith:2006zl}. 
Arguably a minor player in acute bleeding, contact activation could also be important in other wound healing contexts. 
Finally, to make the model more clinically relevant, we should include the biochemical processes responsible for clot formation and clot dissolution (fibrinolysis).
Clot formation is driven by thrombin activity, while fibrinolysis is driven by plasmin activity, a key enzyme that cleaves fibrin (one of the main materials in a clot).
Similar to coagulation, fibrinolysis is managed by several activating and inhibitory factors which control the balance between clot formation and dissolution.
Tissue plasminogen activator (t-PA) and urokinase activate plasmin, along with contact pathway factors such as fXIa. On the other hand, thrombin activatable fibrinolysis inhibitor (TAFI)
inhibits the degradation of fibrin by plasmin. Also, similar to coagulation, there is considerable fibrinolysis and contact pathway data sets that can be used to train the model.
Lastly, the choice of max/min integration rules or the particular form of the transfer functions could be generalized to include other rule types and functions. 
Theoretically, an integration rule is a function whose domain is a set of transfer function inputs, and whose range is $v\in[0,1]$.
Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
Alternative integration rules such as the mean might have different properties which could influence model identification or performance. 
For example, a mean integration rule would be differentiable, which allows derivative-based optimization approaches to be used. 
The particular form of the transfer function could also be explored. We choose a Hill-like function because of its 
prominence in the systems and synthetic biology community. 
However, the only mathematical requirement for a transfer function is that it map a non-negative continuous or categorical variable into the 
range $f\in[0,1]$. Thus, many types of transfer functions are possible.
